Carryover effects in free recall reveal how prior experiences influence memories of new experiences

### 2023-01-13 paper/main.tex
finished complete draft of results section (with all figures and stats)

If participants show different sensitivities to order manipulations, how do
their behaviors carry over to later lists? We found that participants who
showed strong feature clustering on early lists often tended to show strong
feature clustering on late lists (Fig.~\ref{fig:clustering-carryover}A; overall
across participants and conditions: $r(179) = 0.592, p < 0.001$; non-visual
feature conditions: all $r$s $\geq 0.350$, all $p$s $\leq 0.058$; color: $r(29)
= -0.071, p = 0.704$; location: $r(28) = 0.032, p = 0.868$; across conditions:
$r(4) = 0.934, p = 0.006$). Although participants tended to show weaker feature
clustering on late lists (Fig.~\ref{fig:fingerprint-trajectories}) on
\textit{average}, the associations between early and late lists for individual
participants suggests that some influence of early order manipulations may
linger on late lists. We found that participants who exhibited larger carryover
in feature clustering (i.e., continued to show strong feature clustering on
late lists) for the semantic order manipulations (but not other manipulations)
also tended to show a larger improvement in recall
(Fig.~\ref{fig:clustering-carryover}B; overall: $r(179) = 0.378, p < 0.001$;
category: $r(28) = 0.419, p = 0.021$; size: $r(28) = 0.737, p < 0.001$;
non-semantic conditions: all $r$s $\leq 0.252$, all $p$s $\geq 0.179$; across
conditions: $r(4) = 0.773, p = 0.072$) on late lists, relative to early lists.
Participants who exhibited larger carryover in feature clustering also tended
to show stronger temporal clustering on late lists (relative to early lists)
for all but the category condition (Fig.~\ref{fig:clustering-carryover}C;
overall: $r(179) = 0.434, p < 0.001$; category: $r(28) = 0.229, p = 0.223$; all
non-category conditions: all $r$s $\geq 0.448$, all $p$s $\leq 0.012$; across
conditions: $r(4) = 0.598, p = 0.210$). We suggest two potential
interpretations of these findings. First, it is possible that some participants
are more ``maliable'' or ``adaptable'' with respect to how they organize
incoming information. When presented with list of items sorted along
\textit{any} feature dimension, they will simply adopt that feature as a
dominant dimension for organizing those items and subsequent (randomly ordered)
items. This flexibility in memory organization might afford such participants a
memory advantage, explaining their strong recall performance. An alternative
interpretation is that each participant comes into our study with a
``preferred'' way of organizing incoming information. If they happen to be
assigned to an order manipulation condition that matches their preferences,
then they will appear to be ``sensitive'' to the order manipulation and also
exhibit a high degree of carryover in feature clustering from early to late
lists. These participants might demonstrate strong recall performance not
because of their inherantly superior memory abilities, but rather because the
specific condition they were assigned to happened to be especially easy for
them, given their pre-experimental tendancies.

@@

% - the key: individual differences.  some participants are more affected by order manips than others.  for those participants, order manips are sticky and affect recall behavior more.
%   - one subtlety: explain fig 5 and 7 scatterplots (and inlays)-- some *conditions* are overall more affective; pattern is more complicated than can be identified with pairwise comparisons
%   - implication re: individual diffs: adaptive condition-- stabilize should improve performance, destabilize should worsen performance

% JRM: STOPPED HERE...

% - are order effects "sticky"?
%   - when *early* lists are sorted by a given feature, how is memory performance on *late* lists affected (relative to feature rich -- late lists only)?
%   - do some order manipulations matter more than others?  e.g. compare semantic vs. lexicographic vs. visual -- late lists only

%   - compare behavior on early vs. late lists for order manipulation condition
%     - does feature clustering on early lists correlate with recall on early (or late) lists?
%     - does feature clustering on late lists correlate with recall on early (or late) lists?
%     - (ditto, but replace "recall" with "temporal clustering")

% - for feature-rich lists, do order effects matter?
%   - recall + dynamics + organization on order-manipulation conditions vs. feature rich
%   - fingerprint trajectories: how much do fingerprints change over time, are they sensitive to order manipulations?

% - are fingerprints maleable?  how does match between fingerprint + presentation order affect recall performance (adaptive conditions)?

% figure: accuracy by list
Figure~\accuracyByList.

% figure: recall initiation
Figure~\recallInit.

% figure: clustering effects

% figure: correlations between clustering scores
Figure~\clusterCorrs.

% figure: feature clustering vs. accuracy, feature clustering vs. temporal clustering

% figure: fingerprint trajectories

% figure: carryover effects: 
%  - clustering early vs. clustering late
%  - clustering carryover vs. accuracy difference
%  - clustering carryover vs. temporal clustering differences

@@

    \includegraphics[width=0.4 \textwidth]{figures/source/accuracy_adaptive}

        \caption{\textbf{Recall performance (adaptive conditons).} The bars
        display the average probability of recall (taken across words, lists,
        and participants) for lists from each adaptive condition. Error bars
        denote bootstrap-estimated 95\% confidence intervals. For additional
        details about participants' behavior and performance during the
        adaptive conditions, see Figure~\dynamicsAdaptive.}

@@

>>>>

If participants show different sensitivities to order manipulations, how do
their behaviors carry over to later lists? We found that participants who
showed strong feature clustering on early lists often tended to show strong
feature clustering on late lists (Fig.~\ref{fig:clustering-carryover}A; overall
across participants and conditions: $r(179) = 0.592, p < 0.001$; non-visual
feature conditions: all $r$s $\geq 0.350$, all $p$s $\leq 0.058$; color: $r(29)
= -0.071, p = 0.704$; location: $r(28) = 0.032, p = 0.868$; across conditions:
$r(4) = 0.934, p = 0.006$). Although participants tended to show weaker feature
clustering on late lists (Fig.~\ref{fig:fingerprint-trajectories}) on
\textit{average}, the associations between early and late lists for individual
participants suggests that some influence of early order manipulations may
linger on late lists. We found that participants who exhibited larger carryover
in feature clustering (i.e., continued to show strong feature clustering on
late lists) for the semantic order manipulations (but not other manipulations)
also tended to show a larger improvement in recall
(Fig.~\ref{fig:clustering-carryover}B; overall: $r(179) = 0.378, p < 0.001$;
category: $r(28) = 0.419, p = 0.021$; size: $r(28) = 0.737, p < 0.001$;
non-semantic conditions: all $r$s $\leq 0.252$, all $p$s $\geq 0.179$; across
conditions: $r(4) = 0.773, p = 0.072$) on late lists, relative to early lists.
Participants who exhibited larger carryover in feature clustering also tended
to show stronger temporal clustering on late lists (relative to early lists)
for all but the category condition (Fig.~\ref{fig:clustering-carryover}C;
overall: $r(179) = 0.434, p < 0.001$; category: $r(28) = 0.229, p = 0.223$; all
non-category conditions: all $r$s $\geq 0.448$, all $p$s $\leq 0.012$; across
conditions: $r(4) = 0.598, p = 0.210$).

@@

We suggest two potential interpretations of these findings. First, it is
possible that some participants are more ``maliable'' or ``adaptable'' with
respect to how they organize incoming information. When presented with list of
items sorted along \textit{any} feature dimension, they will simply adopt that
feature as a dominant dimension for organizing those items and subsequent
(randomly ordered) items. This flexibility in memory organization might afford
such participants a memory advantage, explaining their strong recall
performance. An alternative interpretation is that each participant comes into
our study with a ``preferred'' way of organizing incoming information. If they
happen to be assigned to an order manipulation condition that matches their
preferences, then they will appear to be ``sensitive'' to the order
manipulation and also exhibit a high degree of carryover in feature clustering
from early to late lists. These participants might demonstrate strong recall
performance not because of their inherantly superior memory abilities, but
rather because the specific condition they were assigned to happened to be
especially easy for them, given their pre-experimental tendancies. To help
distinguish between these interpretations, we designed an \textit{adaptive}
experimental condition (see \textit{Adaptive condition}). The primary
manipulation in the adaptive condition is that participants each experience
three key types of lists. On \textit{random} lists, words are ordered randomly
(as in the feature rich condition). On \textit{stabilize} lists, the
presentation order is adjusted to be maximally similar to the current estimate
of the participant's memory fingerprint (see \textit{Online “fingerprint”
analysis}). Third, on \textit{destabilize} lists, the presentation is adjusted
to be \textit{minimally} similar to the current estimate of the participant's
memory fingerprint (see \textit{Ordering ``stabilize'' and ``destabilize''
lists by an estimated fingerprint}). The orders in which participants
experienced each type of list were counterbalanced across participants to help
reduce the influence of potential list order effects. Because the presentation
orders on stabilize and destabilize lists are adjusted to best match each
participant's (potentially unique) memory fingerprint, the adaptive condition
removes uncertainty about whether participants' assigned conditions might just
``happen'' to match their preferred ways or organizing their memories.

@@

    \includegraphics[width=\textwidth]{figures/adaptive_results}

        \caption{\textbf{Adaptive free recall.} \textbf{A.} Average probability
        of recall (taken across words, lists, and participants) for lists from
        each adaptive condition. \textbf{B.} Average temporal clustering scores
        for lists from each adaptive condition. \textbf{C.} Recall probability
        versus temporal clustering scores by participant (main panel; each
        participant contributes one dot per condition) and averaged within
        condition (inset; each dot represents a single condition). All panels:
        Error bars and ribbons denote bootstrap-estimated 95\% confidence
        intervals. For additional details about participants' behavior and
        performance during the adaptive conditions, see
        Figure~\dynamicsAdaptive.}

@@

Overall, we found that participants tended to be better at remembering words on
stabilize lists relative to words on random ($t(59) = 1.740, p = 0.087$) or
destabilize ($t(59) = 1.714, p = 0.092$) lists (Fig.~\ref{fig:adaptive}A).
Participants showed no reliable differences in their memory performance on
destabilize versus random lists ($t(59) = -0.249, p = 0.804$). Participants
also exhibited stronger temporal clustering on stabilize lists, relative to
random ($t(59) = 3.554, p = 0.001$) and destabilize ($t(59) = 4.045, p <
0.001$) lists (Fig.~\ref{fig:adaptive}B). We found no reliable differences in
temporal clustering for items on random versus destabilize lists ($t(59) =
-0.781, p = 0.438$).

As in the other experimental manipulations, participants in the adaptive
condition exhibited substantial variability with respect to their overall
memory performance and their clustering tendencies (Fig.~\ref{fig:adaptive}C).
We found that individual participants who exhibited strong temporal clustering
scores also tended to recall more items. This held across subjects, aggregating
across all list types ($r(178) = 0.721, p < 0.001$), and for each list type
individually (all $r$s $\geq 0.683$, all $p$s $\leq 0.001$). Taken together,
the results from the adaptive condition suggest that each participant comes
into the experiment with their own unique memory organization tendencies, as
characterized by their memory fingerprint. When participants study lists whose
items come pre-sorted according to their unique preferences, they tend to remember
more and show stronger temporal clustering.
